openTSNE: a modular \proglang{Python} library for t-SNE dimensionality reduction and embedding

### 2023-05-26 2023-05-revision-1/response-to-reviewers.tex

\reviewc{r2q4}{page 7: initialization paragraph only mentions PCA, would it make sense to mention that openTSNE also supports Laplacian eigenmaps as initialization?}

\answer{
We now mention this option in the corresponding paragraph on page 8 and have also added a reference to Laplacian eigenmaps in Section 2.4 in the paragraph on global initialization.
}

\reviewc{r2q5}{It would be interesting to add UMAP to the runtime benchmark. UMAP is a very popular alternative to t-SNE, and is often thought to be much faster, which is not actually the case for openTNSE. Perhaps a separate benchmark figure could focus on openTSNE vs UMAP (depending on the sample size)? This may depend on the used affinity model, as uniform affinity with 15 neighbors should be much faster than the standard affinity model (with 90 neighbors) for large sample sizes. If so, it may be of interest to show both and discuss.}

\answer{
Initially, we decided to omit a comparison to UMAP, as it is a different algorithm and a one-to-one comparison is difficult and not the focal point of the manuscript. However, given the popularity of both algorithms, we agree it makes sense to include a benchmark comparison between their implementations. However, given the often misunderstood differences between the two algorithms, we feel it is necessary to provide a throughout discussion on their benchmark results. We have added a subsection in Section 5.2, comparing the performance of the \texttt{openTSNE} and \texttt{umap-learn} libraries. The uniform affinity model is also an interesting and useful addition in the openTSNE framework, and we have added a comparison to that as well, as it could also be used in real-world scenarios when embedding massive data sets.
}

### 2023-05-29 2023-05-revision-1/main.tex
paper: update benchmarks figures and finalize umap figure

\begin{figure*}[ht]
  \centering
  \includegraphics[width=0.75\textwidth]{benchmarks_python-final}
  \caption{\label{fig:benchmarks_py}
  We benchmark \opentsne (v1.0.0) against three popular open-source
  implementations from \pkg{scikit-learn}~\citep{pedregosa2011scikit}
  (v1.1.2), \pkg{MulticoreTSNE}~\citep{Ulyanov2016} (v0.1), and
  \pkg{FIt-SNE}~\citep{linderman2019fast} (v1.1.0). Experiments were run on a
  on an Intel(R) Xeon(R) CPU E5-1650 equipped with 128 GB of memory. Notice
  that \opentsne scales similarly to \pkg{FIt-SNE}, as they both use the
  same interpolation-based approximation scheme, while \pkg{scikit-learn} and
  \pkg{MulticoreTSNE} utilize the Barnes-Hut approximation.
}
\end{figure*}

@@

\begin{figure*}[ht]
  \centering
  \includegraphics[width=0.75\textwidth]{benchmarks_langs-final}
  %\includegraphics[width=\textwidth]{benchmarks-combined}
  \caption{\label{fig:benchmarks_lang}
  We select the fastest available, open-source implementations of the t-SNE algorithm from the \proglang{Python}, \proglang{R}, and \proglang{Julia} programming languages. We choose \opentsne (v1.0.0) for \proglang{Python}, \pkg{Rtsne}~\citep{krijthe2015rtsne} (v0.15) for \proglang{R}, and \pkg{TSne.jl}~\citep{julia_tsne} (v1.3.0) for \proglang{Julia}. The benchmarks reflect the time complexity of the implemented approximation schemes. For instance, \pkg{TSne.jl} implements a naive $\mathcal{O}(N^2)$ algorithm, making it prohibitively expensive for all but the smallest of data sets. \pkg{Rtsne} scales similarly to \pkg{MulticoreTSNE} in \proglang{Python}, as it implements the $\mathcal{O}(N \log N)$ Barnes-Hut approximation scheme. Interestingly, the multi-threaded implementation version of \pkg{Rtsne} crashes on benchmark data sets containing over 500,000 data points due to a lack of stack memory. We found these crashes to be inconsistent and somewhat random. \opentsne is the only library implementing the more recent FIt-SNE approximation scheme, which makes it suitable for larger data sets. }
\end{figure*}

@@

>>>>

\begin{figure*}[ht]
  \centering
  \includegraphics{benchmarks_python-final}
  \caption{\label{fig:benchmarks_py}
  We benchmark \opentsne (v1.0.0) against three popular open-source
  implementations from \pkg{scikit-learn}~\citep{pedregosa2011scikit}
  (v1.1.2), \pkg{MulticoreTSNE}~\citep{Ulyanov2016} (v0.1), and
  \pkg{FIt-SNE}~\citep{linderman2019fast} (v1.1.0). Experiments were run on a
  on an Intel(R) Xeon(R) CPU E5-1650 equipped with 128 GB of memory. Notice
  that \opentsne scales similarly to \pkg{FIt-SNE}, as they both use the
  same interpolation-based approximation scheme, while \pkg{scikit-learn} and
  \pkg{MulticoreTSNE} utilize the Barnes-Hut approximation.
}
\end{figure*}

@@

\begin{figure*}[ht]
  \centering
  \includegraphics{benchmarks_langs-final}
  %\includegraphics[width=\textwidth]{benchmarks-combined}
  \caption{\label{fig:benchmarks_lang}
  We select the fastest available, open-source implementations of the t-SNE algorithm from the \proglang{Python}, \proglang{R}, and \proglang{Julia} programming languages. We choose \opentsne (v1.0.0) for \proglang{Python}, \pkg{Rtsne}~\citep{krijthe2015rtsne} (v0.15) for \proglang{R}, and \pkg{TSne.jl}~\citep{julia_tsne} (v1.3.0) for \proglang{Julia}. The benchmarks reflect the time complexity of the implemented approximation schemes. For instance, \pkg{TSne.jl} implements a naive $\mathcal{O}(N^2)$ algorithm, making it prohibitively expensive for all but the smallest of data sets. \pkg{Rtsne} scales similarly to \pkg{MulticoreTSNE} in \proglang{Python}, as it implements the $\mathcal{O}(N \log N)$ Barnes-Hut approximation scheme. Interestingly, the multi-threaded implementation version of \pkg{Rtsne} crashes on benchmark data sets containing over 500,000 data points due to a lack of stack memory. We found these crashes to be inconsistent and somewhat random. \opentsne is the only library implementing the more recent FIt-SNE approximation scheme, which makes it suitable for larger data sets. }
\end{figure*}

@@

\subsubsection{Comparison to UMAP}

In this section, we compare the performance of the \opentsne (v1.0.0) and \pkg{umap-learn} (0.5.3) libraries. \pkg{umap-learn} is the official \textsf{Python} implementation of the UMAP algorithm~\citep{2018arXivUMAP}, another dimensionality reduction technique used for the visualization of high-dimensional data. A throughout comparison of the two algorithms is outside the scope of this manuscript, and we here instead highlight only their similarities and differences as they relate to the runtime of the implementations.

Both t-SNE and UMAP are instances of force-directed layout algorithms in which each data point acts as a particle, attracting and repelling other particles. In the first step, both algorithms use the high-dimensional data set to construct a $k$-nearest neighbors graph (KNNG). Then, the low-dimensional embedding is optimized such that data points connected in the KNNG attract one another, and all data points exert some small, repulsive force. The two methods differ in the KNNG construction process, the specification of the attractive and repulsive forces, and their optimization approaches.

The primary factor dictating the runtime of both methods is the number of nearest neighbors used when constructing the KNNG. Apart from the runtime required to find the $k$-nearest neighbors, the number of neighbors directly influences the runtime of each optimization step. In each step, every data point exerts an attractive force on each of its neighbors in the low-dimensional embedding. Therefore, the larger the number of neighbors considered, the longer the runtime. Modern implementations of the t-SNE algorithm construct a KNNG with $3 \cdot u$ nearest neighbors where $u$ is the desired perplexity. By default, \opentsne (v1.0.0) uses a perplexity value of 30, meaning that 90 nearest neighbors need to be found and considered for each data point. In contrast, the \pkg{umap-learn} package (v0.5.3) constructs a KNNG using only 15 neighbors by default, greatly reducing the time spent on nearest neighbor search and point interactions compared to t-SNE.

Alternatively, empirical evidence indicates that replacing the standard perplexity-based Gaussian affinity model in t-SNE with a uniform affinity model produces markedly similar embeddings to that of standard t-SNE. For instance, running standard t-SNE with perplexity $u=30$ constructs a KNNG with $3 \cdot u = 90$ neighbors. However, when using the equivalent uniform affinity model, only $u$ neighbors are needed. This makes the uniform affinity model an attractive option for the visualization of large data sets.

In Fig.~\ref{fig:benchmarks_umap}, we include benchmarks of the standard perplexity-based t-SNE algorithm, its uniform affinity model variant, and UMAP. Differently from the previous benchmarks which were run using older parameter settings to accomodate different t-SNE implementations, we now run both libraries using their default parameter settings to measure the runtime of typical, real-world use-cases.
\begin{figure*}[ht]
  \centering
  \includegraphics{benchmarks_umap-final}
  \caption{\label{fig:benchmarks_umap} We benchmark \opentsne (v1.0.0) against \pkg{umap-learn} (v0.5.3), the official \textsf{Python} implementation of the UMAP algorithm. Notice that the standard single-threaded perplexity-based variant of t-SNE runs slower than UMAP but outperforms UMAP in the multi-threaded case. We also benchmark the uniform affinity model variant of t-SNE, which outperforms both standard t-SNE and UMAP.}
\end{figure*}

Given that, by default, UMAP considers six times fewer neighbors than standard t-SNE, it is unsurprising that the single-threaded variant of UMAP outperforms the single-threaded implementation of standard t-SNE. Suprisingly, however, the single-threaded uniform affinity variant of t-SNE outperforms not only than the single-threaded variant of UMAP, but also its multi-threaded variant, despite dealing with half as many neighbors. This result likely stems from the fact that UMAP is written in pure \textsf{Python}, but uses \pkg{numba}~\citep{lam2015numba} to accelerate computation. On the other hand, critical elements of the \opentsne library are written in \pkg{cython}~\citep{behnel2011cython}, which is subsequently transpiled into \textsf{C++}, which may be able to be more aggresively optimized by the \textsf{C++} compiler. Another contributing factor may be the efficiency of the approximation schemes. While both the UMAP and FIt-SNE algorithms are asymptotically linear in the number of samples, their real-world runtimes may differ significantly.

The multi-threaded benchmarks show that \opentsne outperforms UMAP in both its standard and uniform variants. The reason for this likely lies in the different optimization procedures used by the two algorithms. At each step of the optimization, \opentsne computes all pairwise interactions between data points in the embedding. While a direct calculation of these interactions would result in quadratic time complexity, \opentsne implements efficient approximation schemes that reduce this to linear time complexity (see Sec.~\ref{sec:meth.approx}). Importantly, these approximation schemes are very amenable to parallelization. This is clearly demonstrated by the runtime differences between the single-core and multi-core performance of \opentsne. On the other hand, UMAP bypasses the need to compute all pairwise interactions between data points using negative sampling. This allows UMAP to scale linearly in the number of samples without the need for additional approximation schemes. However, negative sampling is less amenable to paralellization and does not benefit from the availability of additional cores as much as t-SNE.

Lastly, we turn our attention to an often overlooked runtime metric -- the runtime per optimization step. Both t-SNE and UMAP comprise two steps. First, both methods construct a KNNG, which runtime is dominated by $k$-nearest neighbor search. In the second step, a low-dimensional embedding is found via optimization. By default, \opentsne runs optimization for 750 iterations. UMAP, on the other hand, uses a heuristic based on the data set size to determine the number of iterations. For data sets containing fewer than 10,000 data points, UMAP opts for 500 iterations. For larger data sets, however, this is reduced to 200 iterations. Fig.~\ref{fig:benchmarks_umap} indicates that even in the single-threaded scenario, the perplexity-based variant of \opentsne requires significantly less time per iteration compared to \pkg{umap-learn}, as each \opentsne benchmark runs $3.75$ times more iterations than \pkg{umap-learn} despite considering more neighbors at each optimization step.
